---
title: "project1"
author: "Nankoma"
date: "2024-06-05"
output: html_document
---

Loading the required packages
```{r}
library(deSolve)
```

Writing the differential equations.
```{r}
seir_equations <- function(time, variables, parameters) {
  with(as.list(c(variables, parameters)), {
    dS <- -beta * I * S
    dE <-  beta * I * S  - rho * E
    dI <-  rho * E - gamma * I
    dR <-  gamma * I
    return(list(c(dS, dE, dI, dR)))
  })
}
```

Defining some values for the parameters

```{r}
N = 42000001
parameters_values <- c(
  beta  = 0.5/ N, # infectious contact rate (/person/day)
  rho = 1/3.3, # 1/latency period
  gamma = 1/10    # recovery rate (/day)
)
```

Defining initial values for the variables
```{r}
initial_values <- c(
  S = 42000000,  # number of susceptibles at time = 0
  E = 0, # number of exposed individuals at time = 0
  I =   1,  # number of infectious at time = 0
  R =   0   # number of recovered (and immune) at time = 0
)
```

The points in time where to calculate variable values
```{r}
time_values <- seq(0, 200) # days
```

### Numerically solving the SIR model
Using the ode() of `deSolve` package to numerically solve the model
```{r}
seir_values_1 <- ode(
  y = initial_values,
  times = time_values,
  func = seir_equations,
  parms = parameters_values 
)
```

Checking the calculated values
```{r}
seir_values_1 <- as.data.frame(seir_values_1)
seir_values_1
```
Plotting the model
```{r}
with(seir_values_1, {
# plotting the time series of susceptibles:
  plot(time, S, type = "l", col = "blue",
       xlab = "time (days)", ylab = "number of people")
# adding the time series of exposed:
  lines(time, E, col = "orange")
# adding the time series of infectious:
  lines(time, I, col = "red")
# adding the time series of recovered:
  lines(time, R, col = "green")
})

# adding a legend:
legend("right", c("susceptibles", "exposed", "infectious", "recovered"),
       col = c("blue", "orange", "red", "green"), lty = 1, bty = "n")
```

R0 value
```{r}
(42000000 + 1) * parameters_values["beta"] / parameters_values["gamma"]
```

### Writing the simulator
```{r}
seir_1 <- function(beta, rho, gamma, S0, E0, I0, R0, times) {
  require(deSolve) # for the "ode" function
  
# the differential equations:
  seir_equations <- function(time, variables, parameters) {
  with(as.list(c(variables, parameters)), {
    dS <- -beta * I * S
    dE <-  beta * I * S - rho * E
    dI <-  rho * E - gamma * I
    dR <-  gamma * I
    return(list(c(dS, dE, dI, dR)))
  })
  }
  
# the parameters values:
  parameters_values <- c(beta  = beta, rho = rho, gamma = gamma)

# the initial values of variables:
  initial_values <- c(S = S0, E = E0, I = I0, R = R0)
  
# solving
  out <- ode(initial_values, times, seir_equations, parameters_values, maxsteps = 100000)

# returning the output:
  as.data.frame(out)
}
```

```{r}
seir_1(beta  = 0.5/ N, rho = 1/3.3, gamma = 1/10, S0 = 42000000, E0 = 0, I0 = 1, R0 = 0, times = seq(0, 200))
```
### Comparing the model's prediction with data
```{r}
data <- read.csv("data/owid-covid-data.csv")
```


```{r}
ug_data <- data[data$iso_code == "UGA",]
```

```{r}
library(lubridate)
day0 = as.Date(ug_data$date[1])
ug_data["day"] <- as.numeric(difftime(as.POSIXct(as.Date(ug_data$date)), as.POSIXct(day0), units =  "days"))
```

```{r}
uganda <- subset(ug_data, select = c("day", "new_cases"))
head(uganda)
```


```{r}
# Identify outliers        
outliers <- boxplot(uganda$new_cases, plot = FALSE)$out
#[1]  33.9 100.0

# Remove outliers
ug_new<- uganda[!(uganda$new_cases %in% outliers), ]
```

Checking for missing values
```{r}
which(is.na(ug_new))
```

Removing missing values
```{r}
ug_new <- na.omit(ug_new)
```

```{r}
which(is.na(ug_new))
```


```{r}
library("ggplot2")
ggplot(ug_new, mapping = aes(x = day, y = new_cases)) +
  geom_line() + geom_point()
```

Subsetting
```{r}
ug_new1 <- ug_new[200:750,]
```


Plotting the covid data set against the model's predictions
```{r}
with(ug_new1, plot(day, new_cases, pch = 19, col = "red", ylim = c(0, 400)))
predictions <- seir_1(beta  = 0.5/ N, rho = 1/3.3, gamma = 1/10, S0 = 42000000, E0 = 0, I0 = 1, R0 = 0, times = ug_new1$day)
with(predictions, lines(time, I, col = "red"))
```
Function to take in parameters and plot figure as output
```{r}
model_fit <- function(beta, rho, gamma, data, N = 42000001, ...) {
  I0 <- data$new_cases[1] # initial number of infected (from data)
  times <- data$day   # time points (from data)
# model's predictions:
  predictions <- seir_1(beta = beta, rho = rho, gamma = gamma, 
                        S0 = N - I0, E = 0, I0 = I0, R0 = 0,
                        times = times)
  
# plotting the observed prevalences:
  with(data, plot(day, new_cases, ...))
# adding the model-predicted prevalence:
  with(predictions, lines(time, I, col = "red"))
}
```

```{r}
model_fit(beta  = 0.5/ N, rho = 1/3.3, gamma = 1/10, ug_new1, pch = 19, col = "red", ylim = c(0, 400))
```

### Estimating the model's parameters
Using Sum of squares
```{r}
predictions <- seir_1(beta  = 0.5/ N, rho = 1/3.3, gamma = 1/10, S0 = 42000000, E0 = 0, I0 = 1, R0 = 0, times = ug_new1$day)
predictions
```
Sum of Squares
```{r}
sum((predictions$I - ug_new1$new_cases)^2)
```

```{r}
# the observed prevalences:
with(ug_new1, plot(day, new_cases, pch = 19, col = "red", ylim = c(0, 10000000)))
# the model-predicted prevalences:
with(predictions, lines(time, I, col = "red", type = "o"))
# the "errors":
segments(ug_new1$day, ug_new1$new_cases, predictions$time, predictions$I)
```
Function that takes parameters and data set and returns sum of squares
```{r}
ss <- function(beta, rho, gamma, data = ug_new1, N = 42000001) {
  I0 <- data$new_cases[1]
  times <- data$day
  predictions <- seir_1(beta = beta, rho = rho, gamma = gamma,   # parameters
                       S0 = N - I0, E0 = 0, I0 = I0, R0 = 0, # variables' intial values
                       times = times)                # time points
  sum((predictions$I[-1] - data$new_cases[-1])^2)
}
```

```{r}
ss(beta  = 0.5/ N, rho = 1/3.3, gamma = 1/10)
```

Find possible values of beta
```{r}
beta_val <- seq(from = 0, to = 0.5/ N, le = 100)
ss_val <- sapply(beta_val, ss, rho = 1/3.3, gamma = 1/10)
```

Minimum value of the sum of squares
```{r}
min_ss_val <- min(ss_val)
min_ss_val
```

Estimate of infectious contact, beta_val vector that corresponds to the minimum value of the sum of squares
```{r}
beta_hat <- beta_val[ss_val == min_ss_val]
beta_hat
```

Visually
```{r}
plot(beta_val, ss_val, type = "l", lwd = 2,
     xlab = expression(paste("infectious contact rate ", beta)),
     ylab = "sum of squares")
# adding the minimal value of the sum of squares:
abline(h = min_ss_val, lty = 2, col = "grey")
# adding the estimate of beta:
abline(v = beta_hat, lty = 2, col = "grey")
```
Checking for minimum rho
```{r}
rho_val <- seq(from = 0.03, to = 3, le = 100)
ss_val <- sapply(rho_val, function(x) ss(beta_hat, gamma = 1/10, x))
(min_ss_val <- min(ss_val))
```
```{r}
(rho_hat <- rho_val[ss_val == min_ss_val])
```

```{r}
plot(rho_val, ss_val, type = "l", lwd = 2,
     xlab = expression(paste("1/latent period ", rho)),
     ylab = "sum of squares")
abline(h = min_ss_val, lty = 2, col = "grey")
abline(v = rho_hat, lty = 2, col = "grey")
```

Doing the same for recovery rate, gamma
```{r}
gamma_val <- seq(from = 0.01, to = 1, le = 100)
ss_val <- sapply(gamma_val, function(x) ss(beta_hat, rho_hat, x))
(min_ss_val <- min(ss_val))
```

```{r}
(gamma_hat <- gamma_val[ss_val == min_ss_val])
```

```{r}
plot(gamma_val, ss_val, type = "l", lwd = 2,
     xlab = expression(paste("recovery rate ", gamma)),
     ylab = "sum of squares")
abline(h = min_ss_val, lty = 2, col = "grey")
abline(v = gamma_hat, lty = 2, col = "grey")
```
Optimisation
```{r}
ss(beta  = 0.5/ N, rho = 1/3.3, gamma = 1/10)
```
```{r}
ss2 <- function(x) {
  ss(beta = x[1], rho = x[2], gamma = x[3])
}
```

```{r}
ss2(c(0.5/ N, 1/3.3, 1/10))
```
Optimising
```{r}
starting_param_val <- c(2.525252e-09, 0.4, 0.15)
ss_optim <- optim(starting_param_val, ss2)
```

```{r}
ss_optim
```

```{r}
ss_optim$par
```


```{r}
N <- 42000001 # total population size
time_points <- seq(min(ug_new1$day), max(ug_new1$day), le = 100) # vector of time points
I0 <- ug_new1$new_cases[1] # initial number of infected
param_hat <- ss_optim$par # parameters estimates
# model's best predictions:
best_predictions <- seir_1(beta = param_hat[1], rho = param_hat[2], gamma = param_hat[3],S0 = N - I0, E0 = 0, I0 = I0, R0 = 0, time_points)$I
# confidence interval of the best predictions:
cl <- 0.95 # confidence level
cl <- (1 - cl) / 2
lwr <- qnorm(p = cl, mean = best_predictions)
upr <- qnorm(p = 1 - cl, mean = best_predictions)
# layout of the plot:
plot(time_points, time_points, ylim = c(0, max(upr)), type = "n",
     xlab = "time (days)", ylab = "prevalence")
# adding the predictions' confidence interval:
sel <- time_points >= 1 # predictions start from the second data point
polygon(c(time_points[sel], rev(time_points[sel])), c(upr[sel], rev(lwr[sel])),
        border = NA, col = adjustcolor("red", alpha.f = 0.1))
# adding the model's best predictions:
lines(time_points, best_predictions, col = "red")
# adding the observed data:
with(ug_new1, points(day, new_cases, pch = 19, col = "red"))
```
```{r}
N = 42000001
parameters_values2 <- c(
  beta  = 3.724879e-09,
  rho = 4.000076e-01,
  gamma = 1.500024e-01
)
```

```{r}
time_values <- seq(0, 500000) # days
```

```{r}
seir_values_2 <- ode(
  y = initial_values,
  times = time_values,
  func = seir_equations,
  parms = parameters_values2 
)
```


```{r}
seir_values_2 <- as.data.frame(seir_values_2)
seir_values_2
```

```{r}
with(seir_values_2, {
# plotting the time series of susceptibles:
  plot(time, S, type = "l", col = "blue",
       xlab = "time (days)", ylab = "number of people")
# adding the time series of exposed:
  lines(time, E, col = "orange")
# adding the time series of infectious:
  lines(time, I, col = "red")
# adding the time series of recovered:
  lines(time, R, col = "green")
})

# adding a legend:
legend("right", c("susceptibles", "exposed", "infectious", "recovered"),
       col = c("blue", "orange", "red", "green"), lty = 1, bty = "n")
```

